Trauma treatment delay

### 2023-09-18 manuscriptlongversion.Rmd
intro

Within the realm of modern medicine, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients. The definition of medical trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020]. As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological. The one we will be focusing on in this paper is physical trauma, which is defined as a serious injury to the body. It is often divided into two subgroups, blunt force trauma and penetrative trauma. Blunt force trauma happens when an object or force strikes the body, often causing bruising, broken bones or deep cuts. Examples of this could be car crashes, falls or direct blows to the body. Penetrative trauma is when an object pierces the skin or body and creates an open wound. Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].

@@

Trauma is and has long been a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally. This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018]. However, this burden is not evenly distributed between or within countries. Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021]. According to the WHO, about 90% of all trauma related deaths occur in low- and middle-income countries, with injury deaths rates being higher than in high income countries. Even within these LMIC, people of poorer socio-economic status have higher death rates in trauma-related injuries. Problems identified with trauma care within these countries were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021]. In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

@@

Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up. Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response. A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018]. This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group. Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020]. Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012]. During the last three decades, the introduction of trauma systems has lowered the incidence of preventable death from approximately 25% to less than 5%. This is attributed to the improvements in care for acute brain injuries and bleeding control. The incidence of late death because of sepsis and multiple organ failure, possibly a result of better and earlier resuscitation [@Asensio_Trunkey_2008].

@@

Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed]. A cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement and implement corrective actions. This would be a meeting where doctors from different specialities sit down and go through cases where possible OFIs were identified. Examples of opportunities for improvement may include lack of resources and management errors. Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].

>>>>

Within the realm of modern medicine, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients. The definition of medical trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020]. As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological. The one we will be focusing on in this paper is physical trauma, which is defined as a serious injury to the body. It is often divided into two subgroups, blunt force trauma and penetrative trauma. Blunt force trauma happens when an object or force strikes the body, often causing bruising, broken bones or deep cuts. Examples of this could be car crashes, falls or direct blows to the body. Penetrative trauma is when an object pierces the skin or body and creates an open wound. Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020]. Differences exist between the two groups, as blunt trauma patients seem to score higher on ISS scale on arrival, require more resources and be hospitalized for a longer period [@FITCH20191201].

@@

Trauma is and has long been a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally. This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018]. Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].

However, this burden is not evenly distributed between or within countries. Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021]. According to the WHO, about 90% of all trauma related deaths occur in low- and middle-income countries, with injury deaths rates being higher than in high income countries. Even within these LMIC, people of poorer socio-economic status have higher death rates in trauma-related injuries. Problems identified with trauma care within these countries were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021]. In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

@@

Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up. Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response. A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018]. This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group. Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020]. Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012]. During the last three decades, the introduction of trauma systems has lowered the incidence of preventable death from approximately 25% to less than 5%. This is attributed to the improvements in care for acute brain injuries and bleeding control. The incidence of late death because of sepsis and multiple organ failure, possibly a result of better and earlier resuscitation [@Asensio_Trunkey_2008]. Functioning trauma systems saves lives in the hospital, but their work outside of the hospital is just as important. Data collection from injured patients, such as mechanism of injury or mortality, are essential for creating databases that can be used for research. In turn, that research can be used for planning injury prevention programs that target the most common injuries in the right ways, e.g. teen drivers, children, specific occupations etc. These injury prevention programs can be planned on trauma center, organizational or government level[@ACS].

@@

Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed]. A cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement and implement corrective actions. This would be a meeting where doctors from different specialities sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database. Examples of opportunities for improvement may include lack of resources and management errors. One common OFI that this paper will look further into is delayed treatment. Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007]. Although it is such a common OFI, the patient level factors associated with delayed treatment are poorly understood.
